---
title: "Calling episomal ecDNA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calling episomal ecDNA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
suppressPackageStartupMessages(library(EpiTracer))
```

`call_simple_excision()` flags ecDNA amplicons most likely generated by a *simple
excision and re-circularisation* event — an **episome** — rather than by a
complex, breakage-driven process. For each amplicon it inspects the
structural-variant breakpoints at the boundaries and applies the *episome
heuristic*: a boundary duplication (DUP) carrying the highest variant fraction
(`VF`), non-gained flanking segments (consistent with a circle excised from an
otherwise diploid region), and a candidate deletion "excision scar".

## The example data

The package bundles a textbook **episomal EGFR ecDNA** on chr7 (fictional sample
`"EXAMPLE01"`), in the AmpliconArchitect + PURPLE-style `GRanges` the caller
expects. The data are synthetic (no patient data).

```{r data}
names(ex_caller_inputs)
```

## Calling

```{r call}
res <- call_simple_excision(
  ecdna_gr        = ex_caller_inputs$ecdna_gr,
  breakpoints_gr  = ex_caller_inputs$breakpoints_gr,
  cnv_gr          = ex_caller_inputs$cnv_gr,
  cancer_genes_gr = ex_caller_inputs$cancer_genes_gr
)

res[, c("event", "svclass", "VF", "duplication_at_boundary",
        "episomal", "has_excision_scar")]
```

The boundary duplications carry the highest `VF` in the amplicon, the flanks are
diploid, and a deletion scar is present — so every breakpoint of this amplicon is
classified `episomal = "TRUE"` with `has_excision_scar = "TRUE"`. An amplicon
with the *same* boundary DUP but **gained** flanks (i.e. not excised from a
diploid region) is instead returned as `episomal = "FALSE"` — the discrimination
the caller is built for.

## Running without AmpliconArchitect

AmpliconArchitect is optional. Pass `ecdna_gr = NULL` and EpiTracer detects the
focal-amplicon seeds directly from the copy-number segments (via
`detect_amplicon_seeds()`), so it can run from SV + CN alone:

```{r aa-free}
res2 <- call_simple_excision(
  ecdna_gr        = NULL,
  breakpoints_gr  = ex_caller_inputs$breakpoints_gr,
  cnv_gr          = ex_caller_inputs$cnv_gr,
  cancer_genes_gr = ex_caller_inputs$cancer_genes_gr
)
unique(res2$episomal)
```

## Using your own data

The inputs mirror the columns EpiTracer expects from real pipelines: ecDNA
amplicon regions from
[AmpliconArchitect](https://github.com/AmpliconSuite/AmpliconArchitect) and
allele-specific SV / copy-number calls from
[PURPLE](https://github.com/hartwigmedical/hmftools/tree/master/purple) (Hartwig
Medical Foundation pipeline). See `?call_simple_excision` for the required
metadata columns — SV/CN calls from any pipeline can be coerced to the same
shape, and missing columns produce an explicit error.
