---
title: "Calling and visualising episomal ecDNA with EpiTracer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calling and visualising episomal ecDNA with EpiTracer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width = 8,
  fig.height = 4.2,
  dpi = 96
)
suppressPackageStartupMessages(library(EpiTracer))
```

**EpiTracer** detects extrachromosomal circular DNA (ecDNA) amplicons that were
most likely generated by a *simple excision and re-circularisation* event — an
**episome** — and visualises the structural rearrangements underlying focal
amplifications. This vignette runs the two core functions end-to-end on a small,
self-contained example that ships with the package, so you can confirm your
installation works in a few seconds before pointing EpiTracer at real data.

```{r load}
library(EpiTracer)
```

## The example data

The package bundles one toy amplicon — a textbook **episomal EGFR ecDNA** on
chr7 in a fictional sample `"EXAMPLE01"` — in the two input shapes the functions
expect. The data are synthetic (no patient data).

```{r data}
# Inputs for the caller (AmpliconArchitect + PURPLE style GRanges)
names(ex_caller_inputs)

# Inputs for the plotter (PURPLE-style data frames)
names(ex_plot_inputs)
```

## 1. Calling episomal ecDNA

`call_episomal_ecdna()` inspects the structural-variant breakpoints at each
amplicon boundary and applies the *episome heuristic*: a boundary duplication
(DUP) carrying the highest variant fraction, non-gained flanking segments
(consistent with a circle excised from a diploid region), and a candidate
deletion "excision scar".

```{r call}
res <- call_episomal_ecdna(
  ecdna_gr        = ex_caller_inputs$ecdna_gr,
  breakpoints_gr  = ex_caller_inputs$breakpoints_gr,
  cnv_gr          = ex_caller_inputs$cnv_gr,
  cancer_genes_gr = ex_caller_inputs$cancer_genes_gr
)

res[, c("event", "svclass", "VF", "duplication_at_boundary",
        "episomal", "has_excision_scar")]
```

The boundary duplications carry the highest VF in the amplicon, the flanks are
diploid, and a deletion scar is present — so every breakpoint of this amplicon is
classified `episomal = "TRUE"` with `has_excision_scar = "TRUE"`. An amplicon
with the *same* boundary DUP but **gained** flanks (i.e. not excised from a
diploid region) is instead returned as `episomal = "FALSE"` — the discrimination
the caller is built for.

## 2. Visualising the amplicon

`plot_sv_linear()` draws allele-specific copy number, the structural-variant
arc, the karyotype ideogram and the gene label for a locus. Here we focus on the
EGFR amplicon on chr7. (`karyotype` and `gene_coord` are omitted, so the bundled
hg38 references are used.)

```{r plot}
p <- plot_sv_linear(
  sample   = "EXAMPLE01",
  cnv_data = ex_plot_inputs$cnv_data,
  sv_data  = ex_plot_inputs$sv_data,
  wgd_data = ex_plot_inputs$wgd_data,
  chromosome       = "chr7",
  chromosome_range = matrix(c(54e6, 56e6), nrow = 1)
)
p
```

By default the plot is returned as a `ggplot` object and nothing is written to
disk. Pass `outdir = "."` to also save a publication-ready PDF, and supply
`snv_data` to add a stacked panel of the sample's small mutations.

## Using your own data

The example inputs mirror the columns EpiTracer expects from real pipelines:

- **`call_episomal_ecdna()`** — ecDNA amplicon regions from
  [AmpliconArchitect](https://github.com/AmpliconSuite/AmpliconArchitect) and
  allele-specific SV / copy-number calls from
  [PURPLE](https://github.com/hartwigmedical/hmftools/tree/master/purple)
  (Hartwig Medical Foundation pipeline). See `?call_episomal_ecdna` for the
  required metadata columns; other SV/CN callers can be coerced to the same
  shape.
- **`plot_sv_linear()`** — copy-number and SV tables keyed by `sample`; it can
  auto-detect amplified loci or take explicit `loci`. See `?plot_sv_linear`.

```{r session, echo = FALSE}
sessionInfo()
```
