---
title: "Get started with EpiTracer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with EpiTracer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
suppressPackageStartupMessages(library(EpiTracer))
```

**EpiTracer** detects extrachromosomal circular DNA (ecDNA) amplicons that were
most likely generated by a *simple excision and re-circularisation* event — an
**episome** — and visualises the structural rearrangements underlying focal
amplifications, from whole-genome sequencing data.

Three short articles walk through the workflow, each end-to-end on bundled,
synthetic example data (no patient data):

1. **[Calling episomal ecDNA](calling.html)** — `call_episomal_ecdna()`: flag
   amplicons that fit the episome model (with or without AmpliconArchitect seeds).
2. **[Plotting copy number, SVs and kataegis](plotting.html)** —
   `plot_sv_linear()`: a copy-number / structural-variant viewer, with an
   optional stacked SNV (kataegis) panel.
3. **[Reconstructing a focal amplicon wave by wave](reconstruction.html)** —
   `plot_sv_reconstruction()`: rebuild a complex amplicon stratified by read
   support.

## Installation

```r
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("alhafidzhamdan/EpiTracer")
```

## The bundled example data

```{r data}
data(package = "EpiTracer")$results[, "Item"]
```

- `ex_caller_inputs` — inputs for `call_episomal_ecdna()` (an episomal EGFR ecDNA).
- `ex_plot_inputs` / `ex_snv` — inputs for `plot_sv_linear()` (an EGFR amplicon
  and a kataegis SNV set).
- `ex_recon_inputs` — a complex CDK4 amplicon for `plot_sv_reconstruction()`.
