---
title: "Calling rearrangement mechanisms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calling rearrangement mechanisms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4.2, dpi = 96
)
suppressPackageStartupMessages({
  library(EpiTracer)
  library(GenomicRanges)
  library(data.table)
})

## Every example uses a small synthetic locus built behind the scenes, so each
## call is reproducible. Real inputs come from PURPLE / GRIDSS-style copy-number
## and structural-variant calls (see `vignette("epitracer")` for the format).
## `sv_data` is a BEDPE-style data.frame of junctions (used for plotting);
## `sv_to_breakends()` melts it into the paired-breakend `breakpoints_gr` the
## callers expect. `cnv_data` is the copy-number segmentation.
cancer_genes_gr <- GRanges("chr7", IRanges(55019017, 55211628), gene = "EGFR")

sv_to_breakends <- function(sv) {
  sv <- as.data.table(sv); sv[, event := paste0("E", .I)]
  a <- sv[, .(seqnames = paste0("chr", chrom1), start = start1, bp_strand = strand1)]
  b <- sv[, .(seqnames = paste0("chr", chrom2), start = start2, bp_strand = strand2)]
  m <- sv[, .(event, svclass, VF, PURPLE_JCN = JCN, PURPLE_CN)]
  d <- rbind(cbind(a, m), cbind(b, m))
  d[, `:=`(end = start, WGS_ID = "S1", PURPLE_AF = 0.9, insLen = 0L, HOMLEN = 0L)]
  makeGRangesFromDataFrame(d, keep.extra.columns = TRUE)
}
cnv_to_gr <- function(cnv) makeGRangesFromDataFrame(as.data.frame(cnv), keep.extra.columns = TRUE)
```

`call_simple_excision()` asks *"is this a simple-excision circle (an episome)?"*.
The **mechanism callers** ask the complementary question — *by which route did a
focal amplification, or a complex rearrangement, form?* Each one takes the same
inputs as the episomal caller (`breakpoints_gr`, `cnv_gr`, `cancer_genes_gr`, and
an optional `ecdna_gr`) and returns those breakpoints annotated with **one
mechanism's flag**.

The mechanisms are not all amplifications: chromothripsis-within-an-ecDNA,
micronucleation and breakage–fusion–bridge produce amplified segments, whereas
breakage–replication/fusion and chromoplexy are balanced rearrangement
signatures that need not raise copy number. The callers are also **independent** —
an amplicon can carry more than one flag — so the workflow is to run the relevant
callers on the same inputs and join their outputs by `WGS_ID` + `ID`.

Each section below shows the **call** and a **plot** of the underlying event; the
inputs are constructed behind the scenes.

## Chromothripsis within an ecDNA

Shattering and random rejoining inside an amplicon. `call_chromothripsis()`
scores ShatterSeek-style hallmarks over the amplified footprint: enough internal
breakpoints, the four junction orientations occurring about equally, and an
**oscillating** copy-number profile.

```{r ct-data, include = FALSE}
lo <- 55000000; hi <- 55500000
sv_data <- data.frame(chrom1 = "7", start1 = lo, chrom2 = "7", start2 = hi,
  strand1 = "-", strand2 = "+", svclass = "DUP", VF = 1000, JCN = 40,
  PURPLE_CN = 40, sample = "S1", stringsAsFactors = FALSE)
cls <- rep(c("DEL", "DUP", "h2hINV", "t2tINV"), each = 3)
st  <- list(DEL = c("+","-"), DUP = c("-","+"), h2hINV = c("+","+"), t2tINV = c("-","-"))
for (k in seq_along(cls)) { a <- lo + round((hi - lo) * k / 13); s <- st[[cls[k]]]
  sv_data <- rbind(sv_data, data.frame(chrom1 = "7", start1 = a, chrom2 = "7",
    start2 = a + 30000, strand1 = s[1], strand2 = s[2], svclass = cls[k], VF = 500,
    JCN = 20, PURPLE_CN = 40, sample = "S1", stringsAsFactors = FALSE)) }
osc <- c(lo, lo + 1e5, lo + 2e5, lo + 3e5, lo + 4e5)
cnv_data <- data.frame(sample = "S1", seqnames = "chr7", start = c(1, 4e7, osc, hi + 1),
  end = c(4e7 - 1, lo - 1, c(osc[-1] - 1, hi), 1.59e8), copyNumber = c(2, 2, 40, 18, 42, 16, 40, 2),
  ploidy = 2, majorAlleleCopyNumber = c(1, 1, 39, 17, 41, 15, 39, 1), minorAlleleCopyNumber = 1)
ecdna_gr       <- GRanges("chr7", IRanges(lo, hi), ID = "S1_amp1", WGS_ID = "S1")
breakpoints_gr <- sv_to_breakends(sv_data)
cnv_gr         <- cnv_to_gr(cnv_data)
```

```{r ct-call}
res <- call_chromothripsis(ecdna_gr, breakpoints_gr, cnv_gr, cancer_genes_gr)
unique(res[, c("chromothripsis", "chromothripsis_conf")])
```

```{r ct-plot}
plot_sv_linear(sample = "S1", cnv_data = cnv_data, sv_data = sv_data,
               chromosome = "chr7", chromosome_range = matrix(c(54.5e6, 56e6), nrow = 1))
```

The oscillating copy number and the fan of differently-oriented internal
junctions are the shattering signature; a clean episome (only a boundary
junction) returns `chromothripsis = "FALSE"`.

## Micronucleation — two ecDNAs fused in a micronucleus

`call_micronucleation()` flags an amplicon joined to *another amplified locus on
a non-homologous chromosome* by a high-VF interchromosomal translocation. Fusing
two different chromosomes into one amplicon is only possible if both were
co-encapsulated — the signature of two episomal ecDNAs recombined in a
micronucleus after shattering.

```{r mn-data, include = FALSE}
sv_data <- data.frame(chrom1 = "7", start1 = 55200000, chrom2 = "12", start2 = 57200000,
  strand1 = "+", strand2 = "-", svclass = "TRA", VF = 3000, JCN = 50, PURPLE_CN = 50,
  sample = "S1", stringsAsFactors = FALSE)
cnv_data <- data.frame(sample = "S1",
  seqnames = c("chr7", "chr7", "chr7", "chr12", "chr12", "chr12"),
  start = c(1, 55e6, 55500001, 1, 57e6, 57300001),
  end   = c(54999999, 55500000, 7e7, 56999999, 57300000, 7e7),
  copyNumber = c(2, 50, 2, 2, 50, 2), ploidy = 2,
  majorAlleleCopyNumber = c(1, 49, 1, 1, 49, 1), minorAlleleCopyNumber = 1)
ecdna_gr       <- GRanges("chr7", IRanges(55e6, 55500000), ID = "S1_amp1", WGS_ID = "S1")
breakpoints_gr <- sv_to_breakends(sv_data)
cnv_gr         <- cnv_to_gr(cnv_data)
```

```{r mn-call}
res <- call_micronucleation(ecdna_gr, breakpoints_gr, cnv_gr, cancer_genes_gr)
unique(res$micronucleation)
```

A genome-wide view (`plot_sv_circos()`) shows the two amplified loci on chr7 and
chr12 joined by the translocation arc:

```{r mn-plot, fig.width = 6, fig.height = 6}
plot_sv_circos(sample = "S1", sv_data = sv_data, cnv_data = cnv_data)
```

## Breakage–fusion–bridge (BFB)

Classical BFB is **telomeric**: after telomere loss, sister chromatids fuse and
the dicentric bridge breaks unevenly each cycle, building a fold-back amplicon
that is *terminal on one arm*, carries a *distal deletion to the telomere*, and
shows a copy-number **staircase**. `call_bfb()` requires all three, and needs
centromere / chromosome-length references to locate the telomere.

```{r bfb-data, include = FALSE}
lo <- 10e6; up <- 20e6; w <- up - lo
sv_data <- data.frame(chrom1 = "7", start1 = c(lo + 0.2 * w, lo + 0.8 * w), chrom2 = "7",
  start2 = c(lo + 0.2 * w + 5e4, lo + 0.8 * w + 5e4), strand1 = "+", strand2 = "+",
  svclass = "h2hINV", VF = 500, JCN = 30, PURPLE_CN = 40, sample = "S1", stringsAsFactors = FALSE)
cnv_data <- data.frame(sample = "S1", seqnames = "chr7",
  start = c(1, lo, lo + w/3 + 1, lo + 2*w/3 + 1, up + 1),
  end   = c(lo - 1, lo + w/3, lo + 2*w/3, up, 1.59e8),
  copyNumber = c(1, 40, 25, 12, 2), ploidy = 2,
  majorAlleleCopyNumber = c(0, 39, 24, 11, 1), minorAlleleCopyNumber = c(0, 1, 1, 1, 1))
ecdna_gr       <- GRanges("chr7", IRanges(lo, up), ID = "S1_amp1", WGS_ID = "S1")
breakpoints_gr <- sv_to_breakends(sv_data)
cnv_gr         <- cnv_to_gr(cnv_data)
```

```{r bfb-call}
res <- call_bfb(ecdna_gr, breakpoints_gr, cnv_gr, cancer_genes_gr,
                centromeres = load_centromeres("hg38"),
                chrom_lengths = load_chrom_lengths("hg38"))
unique(res[, c("bfb", "bfb_anchor", "n_foldbacks")])
```

```{r bfb-plot}
plot_sv_linear(sample = "S1", cnv_data = cnv_data, sv_data = sv_data,
               chromosome = "chr7", chromosome_range = matrix(c(1, 25e6), nrow = 1))
```

The distal deletion to the p-telomere and the descending copy-number staircase
are the BFB signature. Drop any one hallmark — a diploid terminus, a flat
plateau, or clustered fold-backs — and the call reverts to `"FALSE"`.

## Breakage–replication/fusion (BRF)

BRF (Mendez-Dorantes / Zhang / Pellman, *Nat Genet* 2026) leaves **adjacent
parallel breakpoints** — two same-orientation breakends, close together, from
*distinct* junctions. Unlike BFB it is not telomere-gated and need not build a
staircase.

```{r brf-data, include = FALSE}
sv_data <- data.frame(chrom1 = "7", start1 = c(55100000, 55110000), chrom2 = "7",
  start2 = c(55400000, 55600000), strand1 = "+", strand2 = "+", svclass = "h2hINV",
  VF = 500, JCN = 30, PURPLE_CN = 50, sample = "S1", stringsAsFactors = FALSE)
cnv_data <- data.frame(sample = "S1", seqnames = "chr7", start = c(1, 55e6, 55500001),
  end = c(54999999, 55500000, 7e7), copyNumber = c(2, 50, 2), ploidy = 2,
  majorAlleleCopyNumber = c(1, 49, 1), minorAlleleCopyNumber = 1)
ecdna_gr       <- GRanges("chr7", IRanges(55e6, 55500000), ID = "S1_amp1", WGS_ID = "S1")
breakpoints_gr <- sv_to_breakends(sv_data)
cnv_gr         <- cnv_to_gr(cnv_data)
```

```{r brf-call}
res <- call_brf(ecdna_gr, breakpoints_gr, cnv_gr, cancer_genes_gr)
unique(res[, c("brf", "n_parallel_pairs")])
```

```{r brf-plot}
plot_sv_linear(sample = "S1", cnv_data = cnv_data, sv_data = sv_data,
               chromosome = "chr7", chromosome_range = matrix(c(54.5e6, 56e6), nrow = 1))
```

## Chromoplexy

Chromoplexy (Baca *et al.* 2013) is a **closed, balanced** chain of
rearrangements cycling through several chromosomes back to its start, at *low*
copy number. It is genome-wide rather than amplicon-anchored, so
`call_chromoplexy()` takes just `breakpoints_gr` and `cnv_gr` and returns **one
row per chain**.

```{r cp-data, include = FALSE}
sv_data <- data.frame(
  chrom1 = c("1", "5", "12"), start1 = c(10000000, 20005000, 30005000),
  chrom2 = c("5", "12", "1"), start2 = c(20000000, 30000000, 10005000),
  strand1 = "+", strand2 = "-", svclass = "TRA", VF = 30, JCN = 2, PURPLE_CN = 2,
  sample = "S1", stringsAsFactors = FALSE)
cnv_data <- data.frame(sample = "S1", seqnames = c("chr1", "chr5", "chr12"),
  start = 1, end = 2e8, copyNumber = 2, ploidy = 2,
  majorAlleleCopyNumber = 1, minorAlleleCopyNumber = 1)
breakpoints_gr <- sv_to_breakends(sv_data)
cnv_gr         <- cnv_to_gr(cnv_data)
```

```{r cp-call}
res <- call_chromoplexy(breakpoints_gr, cnv_gr)
res[, c("topology", "n_junctions", "n_chromosomes", "chromosomes")]
```

```{r cp-plot, fig.width = 6, fig.height = 6}
plot_sv_circos(sample = "S1", sv_data = sv_data, cnv_data = cnv_data)
```

The closed chr1 → chr5 → chr12 → chr1 cycle at diploid copy number is the
chromoplexy signature. An **open** chain, or an **amplified** cycle (an amplicon,
not a balanced event), is not called.

## Combining the callers

Because each caller reports one signature and leaves the others untouched, run
the relevant callers on the same inputs and join their per-breakpoint outputs by
`WGS_ID` + `ID`. An amplicon then carries an explicit flag for every mechanism
tested — `episomal`, `chromothripsis`, `micronucleation`, `bfb`, `brf` — rather
than being forced into a single class. See `?call_simple_excision` for the shared
input contract, and `vignette("reconstruction")` for reading the founder junction
off a VF-stratified reconstruction.
