---
title: "Reconstructing a focal amplicon wave by wave"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reconstructing a focal amplicon wave by wave}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width = 7,
  fig.height = 8.5,
  dpi = 96
)
suppressPackageStartupMessages(library(EpiTracer))
```

`plot_sv_reconstruction()` rebuilds a focal amplicon **wave by wave**. It
stratifies the structural-variant junctions by read support (`VF`) and draws one
panel per stratum, from the highest-`VF` **founder** junction down to the
lowest-`VF` late additions — so you can read the order in which the amplicon was
assembled. (Requires the \pkg{patchwork} package.)

## The example data

The bundled `ex_recon_inputs` is a synthetic EGFR amplicon on chr7 carrying many
duplication junctions spanning read support from `VF` 6 to 520:

```{r data}
str(ex_recon_inputs$sv_data)
```

## Reconstruction

```{r recon}
plot_sv_reconstruction(
  sample   = "EXAMPLE01",
  cnv_data = ex_recon_inputs$cnv_data,
  sv_data  = ex_recon_inputs$sv_data,
  wgd_data = ex_recon_inputs$wgd_data,
  chromosome = "chr7"
)
```

Reading it top to bottom: the first panel isolates the single highest-`VF`
**founder** junction (the earliest, highest-copy event); each subsequent panel
folds in the next-lower `VF` stratum, and the amplified region grows as later
junctions are added. Read support is shown on the right-hand axis, and prior
(higher-`VF`) junctions are drawn faintly for context.

## Controls

By default the number of strata is chosen automatically (`k = "auto"`). Common
adjustments:

- **`k`** — fix the number of read-support strata.
- **`vf_breaks`** — explicit `VF` cut points instead of automatic clustering.
- **`min_vf`** — drop low-support junctions before clustering.
- **`isolate_founder = FALSE`** — keep the single top junction inside its
  cluster rather than in its own founder panel.

See `?plot_sv_reconstruction` for the full set of controls, and the
[get-started vignette](epitracer.html) for the copy-number / SV viewer
`plot_sv_linear()`.
